---
title: "Modeling Carbon Dioxide Emissions at Mauna Loa, Hawaii"
subtitle: "ECO 5740 - Forecasting and Time Series Models (Spring 2025)"
author: "Ethan W. Turner"
date: "`r Sys.Date()`"
execute:
  echo: true
  warning: false
  message: false
  cache: true
format:
  pdf:
    self-contained: true
number-sections: true
toc: true
toc-depth: 3
knitr:
  opts_chunk:
    eval: true # Enable code execution globally
    chunk_output_type: inline  # Display code output inline, avoiding overflow
editor: 
  markdown: 
    wrap: sentence
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(fpp3) 
library(readr) # for reading csv files
library(readxl) # for reading excel files

# Please refrain from using any other libraries for this project unless it is absolutely required. 

library(latex2exp)
#To apply an example from class

library(tsibble) 
#to use the function yearmonth()

library(fable.prophet)
#to run given code

library(urca)


# To generate a PDF document from the .qmd file, you need to have MiKTeX installed on your computer. MiKTeX is a LaTeX distribution that Quarto uses to compile the document into a PDF. You can install MikTex from https://miktex.org/download.

# To run the R code within the R chunk in your Quarto (.qmd) file, you need to enable code evaluation. Change the value of eval from false to true.
# knitr:
#   opts_chunk:
#     eval: true
```

# Abstract

Using monthly average data available from the Mauna Loa observatory via NOAA, this forecasting analysis concludes that atmospheric carbon dioxide levels will continue to increase. Our final model, an ARIMA selection chosen from a shortlist of simple and advanced singular and ensemble models, projects a higher rate of increase than seen on average in the first 66 years of the observatory's existence, culminating in new record levels of CO2 every year and surpassing 450 ppm (parts per million) by 2035, dwarfing the current record peak of 426.91 ppm.


# Introduction

1.  Introduction and Motivation: Briefly introduce your variable of interest and its significance.
    Explain how forecasting this variable benefits key stakeholders (e.g., policymakers, businesses).
    State the problem or question your forecast addresses.

Carbon dioxide possesses significance as the foremost greenhouse gas in the atmosphere and one of the major harbingers of climate change. Understanding that the gas contributes to global warming and therefore icecap melt, wildlife disruption, and other secondary effects has led to the acceptance of carbon dioxide (CO2) as a crucial metric for forecasting. Thus, the rate of CO2's increase in the atmosphere can directly inform concerning the speed at which the climate may be expected to change.

Many stakeholders benefit from forecasting of CO2. Governmental policymakers require the data to set targets and plan transitions to net-zero CO2 plans, perhaps even generating multiple long-term forecasts based on possible industry limits and policy changes. At the industry level, carbon dioxide plays an informative role by relating to aspects of climate change that have massive ramifications on industry practices. For example, in food production climate change has shifted ideal crop patterns. Understanding the rate and future of these changes will aid food producers in setting the pace of their necessary transitions. Likewise, property insurance companies require projections to set reasonable insurance rates on coastal properties that will be at risk with rising sea levels linked to climate change. Setting the correct rate requires an understanding of the level of risk associated with a location in the near and distant future. Thus, the benefits of forecasting carbon dioxide are enormous, and this forecast seeks to address both shorter and longer-term patterns in atmospheric CO2 levels that will help these stakeholders plan effectively for the future.



2.  Data Relevance: Describe your dataset, including its source, timeframe, and key attributes.
    Justify why this dataset is appropriate for your analysis.
    
    
This report utilizes monthly data from March 1958 to December 2024 on carbon dioxide (CO2) readings from Mauna Loa Observatory, Hawaii. The observatory falls under the umbrella of the National Oceanic and Atmospheric Association (NOAA), and its readings are available on the NOAA website at https://gml.noaa.gov/ccgg/trends/data.html. Specifically, this forecast both uses and projects values of the monthly average readings from Mauna Loa.

The data is complete without any missing monthly averages, despite several minor complications. At the observatory, CO2 is measured daily, but there have been several gaps since 1958 due to natural disasters or other obstacles that prevented readings from being taken. Insofar as each month possesses a monthly average in the dataset, this does not create missing values, but does mean that certain days in months having disrupted days contribute more to a specific monthly average than more complete months.

This dataset commands great respect in the field of climate change studies, and demonstrates far more reliability than a run-of-the-mill series of measurements from the average outpost. Mauna Loa provides an uncharacteristically long record of serial CO2 measurements from the same location, especially considering that the advent of this dataset in 1958 occurred before the popularization of climate change analysis and the scientific consensus on man-made climate change's existence. Charles David Keeling, the key figure in the formation of the Mauna Loa Observatory, is credited as bringing CO2 levels to wider awareness through the Mauna Loa data and confirming the rapid increase of atmospheric CO2 across the first decades of the observatory's records. Thus, Mauna Loa's data is reputable and carries great weight in the scientific community, marking it as an ideal dataset to forecast atmospheric CO2 levels.   
  

3.  Scientific Context: Reference relevant studies or credible sources to highlight the importance of forecasting your variable.
    Mention trends or findings that make the variable interesting or challenging to predict.



The United Kingdom Met Office also measures CO2 levels from the Mauna Loa Observatory as a symbol of global atmospheric levels and the implications of any changes. However, their methods contain more contextual variables such as the prominence of the El Niño climate pattern in a particular year in raising the sea surface temperatures (Betts et al., 2025). Our forecast relies solely on past CO2 monthly averages to predict the future, but issues with the Met Office forecast's reliability implies that a simple method may surpass a more complex canvassing of secondary variables. Looking at the relation of the Met Office's forecast for annual mean CO2 levels compared to the true value, the Met Office has failed to capture the true annual mean in its margin of error for the past two years (2023 and 2024). Thus, this forecast hopes to present greater accuracy purely through time series methods, since CO2 levels are proving volatile respective to official, well-research predictions.

In addition, the Met Office's research underlines a crucial role of forecasting CO2 levels. With these forecasts, researchers not only project CO2 but corresponding imminent increases in world temperature. For instance, the Met Office linked their most recent forecast to the Intergovernmental Panel on Climate Change's (IPCC) scenarios where global temperatures only increase 1.5°C relative to current levels. This goal represents a target of minimum environmental destruction, but the Met Office noted that the level of forecasted CO2 increase is inconsistent with the IPCC's goal. Thus, CO2 forecasts carry wide implications in the climate change research field.  





4.  Key Findings and Stakeholder Benefits: How will the variable behave in the next h periods (increase, decrease, or remain stable)?
    Why is this behavior significant?
    Identify stakeholders who benefit from your forecasts.
    Explain how forecasts help them make decisions, reduce risks, or seize opportunities.


In the next decade, this forecast estimates that carbon dioxide levels will increase at a more rapid rate than in previous decades. This carries significance by mandating even swifter changes in how companies and governments react to shifts in the climate. While climate change and its accompanying natural disasters and forced adaptations bring expenses onto private and public institutions, the larger rate of increase in CO2 levels will help several industries. For example, shipping companies will possess access to a greater amount of shipping routes near the Arctic region due to the melting of ice. This increases efficiency in marine transportation. 

However, more industries benefit from a knowledge of the projected levels of future increase rather than from the increase itself. Winemakers, for example, face a change in optimal winemaking regions due to climate zones shifting amidst increasing temperatures. Anticipating a more rapid increase in temperatures will help winemaking businesses stay ahead of disruptive changes, allowing them to invest in areas on the edge of becoming hospitable for winemaking. Thus, businesses may reap the benefits as this area rapidly becomes more suitable. Although this requires initial expense due to abandoning areas worsening due to climate change, adaptation informed by atmospheric CO2 levels lessens the economic blow suffered in industries like these.




# Data Description and Sources

In total, there are 802 monthly averages from March 1958 to December 2024, each representing the monthly average CO2 reading. Although many individual days do not possess readings, each month retains its calculated mean through all recorded values. In the dataset, year and month are recorded separately. In addition, the dataset records the passage of time as a single decimal value rising incrementally. For instance, the reading of March 1958 is expressed as "1958.203". The mean amount of CO2 present in the air each day over the course of a month appears in the dataset in units of parts per million (ppm). Starting in 1974, researchers also began recording standard deviation and uncertainty across the month's readings. Until then, these columns are marked by negative numbers to avoid confusion with true records.


# Statistical Methods

## Time Series Graphics

### Set up data as a time series object

```{r, eval = TRUE}
# Bring my csv file from download to R and call it mydata
mydata <- read_csv("C:/Users/ethan/Downloads/co2_mm_mlo.csv", skip = 40)
# set up data as a time series object

mydata <- mydata |>
  mutate(date = paste(year, month, sep = "-")) |> #I needed to combine columns because the "decimal date" column was not producing the correct dates with the recommended functions
  mutate(date = yearmonth(date)) %>% # it is monthly data
  as_tsibble(index = date)
```

### Plot the data

```{r, eval=TRUE}
mydata |>
  autoplot(average) +
  ggtitle("Mauna Loa CO2 Readings: 03/1958 to 12/2024") + 
  xlab("Date") + 
  ylab("CO2 (ppm)")
```

### Time series patterns

```{r}
# Calculate and plot the ACF for the CO2 variable
mydata %>%
  ACF(average, lag_max = 60) %>% 
  autoplot() +
  ggtitle("Mauna Loa CO2 Monthly Means ACF: 03/1958 to 12/2024")
```

```{r}
mydata |>
  gg_season(average) +
  labs(title = "Seasonal Plot: Mauna Loa CO2 Monthly Means",
       y = "CO2 (ppm)",
       x = "Time (Month)")
```

```{r}
mydata |>
  gg_subseries(average) +
  labs(title = "Mauna Loa CO2 Readings: Monthly Means",
    y = "CO2 average (ppm)"
  )

```

The ACF plot shows slight seasonality, with peaks and troughs every 12 months.
However, the most clear component of the visual shows a slow decay of autocorrelation values, indicating a trend.
The seasonal plot, the next one down, displays a general increase of monthly mean CO2 readings from below 325 ppm to approximately 425 ppm over the years, as well as annual peaks around May and June.
Finally, the seasonal subseries plot leads to the same conclusions, showing sharp increases over time in CO2 levels as well as some seasonality with a peak around May and a low point around September.

The data lacks greatly exceptional shifts in trend or seasonality over time.
There are no clear outliers, though the data's trend exhibits a slightly exponential quality in its increase.
There do not appear to be any unusual instances where the trend ceased or reversed, as in the seasonal plot each color-coded time period every individual grouped year appears with adjacent years.
Although the first raw plot of the data exhibits some variation in in the smoothness of seasonal transitions with the degree and constancy of increase and decrease, the seasonal trend always remains intact.

Contextually, these visuals underline the concerning rise in CO2 levels over the past few decades, emblematic of a currently losing battle against climate change.
The overwhelming positive trend dwarfs the contrastingly slight seasonality fluctuations.
Concerningly, the trend shows no signs of diminished progress in more recent years, indicating that any policy changes or increased awareness of CO2 emissions have impacted atmospheric CO2 levels.
This project will help to forecast the future increases of CO2 while maintaining the status quo of insufficient abatement.

## Transformations and Adjustments

Of the four major types of adjustment, three definitely do not apply. Because our data involves merely monthly averages and does not delve into financial or population figures, we have no use for calendar, population, or inflation adjustments. In the specific case of a calendar adjustment pertaining to this project, months such as February are only different from longer months due to containing fewer days to extract a monthly CO2 average. Because our data does not involve totals, a calendar adjustment does not apply here. The elimination of the other two transformation types, given the nature of our data, is self-explanatory.

However, I will employ a Box-Cox transformation with the Guerrero method to stabilize the variance across the data, especially after noting the software indicates a value of *lambda* for which the variance is more stable.

```{r}
lambda <- mydata |> # save the data
  features(average, features = guerrero) |> # extract "average" from the "mydata" and calculate guerrero feature (lambda).
  pull(lambda_guerrero) # pull the value for lambda
```

Our value of *lambda* is approximately 0.91. In the Box-Cox transformation, our transformed data defined as w~t~ at each interval t will thus take the form (y~t~^0.91^ -1) / 0.91. Via software, we choose this value to homogenize the seasonal variation across the entire series, which will aid our forecasting.

```{r}
#First, plot the original time plot for a frame of reference
mydata |> 
  autoplot(average) +
  ggtitle("Mauna Loa CO2 Readings: 03/1958 to 12/2024") + 
  xlab("Date") + 
  ylab("CO2 (ppm)")

# use the above calculated lambda and plot the "average" data
mydata |>
  autoplot(box_cox(average, lambda)) + 
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed CO2 monthly average with lambda = ",
         round(lambda,2))))

#input a column with the transformed values while still preserving the original values
mydata$trans_average <- box_cox(mydata$average, lambda)

```

The data's time plot does not appear to have been greatly affected by the transformation, as evidenced by a side-by-side comparison with the original. The only major change has occurred within the y-axis scaling, with our transformed data covering a span of approximately 200 to 260 while the original covered around 310 to 430. Despite this scale change, the plot still carries an evident positive trend and seasonality, while the few uncharacteristically jagged transitions between seasonal phases remain largely unimpacted by the transformation.

Although the transformation targets homogenizing seasonal variation, the shift between the original and new time plot is very subtle. In the data's later years, the transformation slightly lessened the magnitude of the seasonal impact. Despite this, our transformed data remains very similar to the original. 

## Time Series Decomposition

This project will now utilize STL (seasonal-trend using the Loess method) decomposition and analyze graphical representations of the results after additively separating the components.

```{r}
#enact STL decomposition

dcmp <- mydata |>
  model(stl = STL(trans_average))

```

```{r}
#view the graph of the numerical results of STL decomposition

components(dcmp) |> 
  autoplot()
```

```{r}
components(dcmp) |> # use the above decomposed data.
  as_tsibble() |> # convert it into a tsibble object.
  autoplot(trans_average, # plot the transformed "average" data.
           colour="gray") + # use gray color for the transformed data.
  geom_line(aes(y = trend), # plot the trend column on y axis in orange (#D55E00) color. 
            colour = "#D55E00") +
  labs(y = "CO2 PPM (transformed)",
       title = "CO2 Emissions at Mauna Loa with Trend Line")
```

```{r}
components(dcmp) |> # use the above decomposed data.
  as_tsibble() |> # convert it into a tsibble object.
  autoplot(trans_average, # plot the transformed "average" data in gray color
           colour = "gray") +
  geom_line(aes(y=season_adjust),  # plot the season_adjust column on y axis in blue (#0072B2) color.
            colour = "#0072B2") +
  labs(y = "CO2 PPM (transformed)",
       title = "CO2 Emissions at Mauna Loa with Seasonal Adjustment")
```

After using STL decomposition, we have successfully extracted trend, seasonal, and remainder components. Unsurprisingly given the clear upwards trajectory of the raw data, the trend is positive and mildly exponential. Extracting the trend reveals several rougher edges, mostly before the year 2000, where the rate of growth appears to have changed abruptly without completely ceasing.

The graphical representation of the seasonal component after the transformation reveals a fluctuation of increase and decrease in magnitude.
It generally shifts between 1 and 2 units each year.
The seasonally adjusted time plot of the transformed data successfully removes most of the data's seasonality to reveal an incessant rise of CO2 levels (albeit not in their original form). Due to some clear remaining fluctuations in the 1960s and 1970s the seasonal adjustment is still not monotonically increasing.

From first glance, the remainder component possesses characteristics associated with white noise, namely a lack of correlation and trend.
There is one curious positive spike in April 2016 where the remainder crosses an absolute value of 0.5 for the first time, but there does not initially appear enough evidence to make a worthwhile conclusion based on one unusual observation.

## Forecasting Methods

### Training set

```{r, eval=TRUE}
# extract the training set to fit the data
training_set <- mydata |>
  filter_index("1958-03" ~ "2011-08") 

# extract the test set  
test_set <- mydata |>
  filter_index("2011-09" ~ "2024-12") 

# set the forecast horizon equals to the test set
h <- nrow(test_set)
```

We desire an 80-20 split between the training and test set. There are 802 total observations in the overall dataset. 20% of 802 is 641.6, which rounds up to 642. With each individual observation representing a monthly CO2 average, 642 observations equates to 53 full years and an additional 6 months. Considering our data begins in March 1958, our training set extends from here to August 2011. Our remaining 160 observations comprise the test set, which starts in September 2011 and extends to December 2024; a total of 13 years and 4 months. 

### Fit the model

```{r, eval=TRUE}
# Fit the models
my_fit <- training_set |>
  model(
    Mean = MEAN(trans_average),
    `Naïve` = NAIVE(trans_average),
    Drift = NAIVE(trans_average ~ drift()),
    `Seasonal naïve` = SNAIVE(trans_average)
  )
```

### Compare the forecast accuracy across models

```{r, eval=TRUE}
# Fit the models
my_fit <- training_set |>
  model(
    "Holt-Winters" = ETS(box_cox(average, lambda) ~ error("A") + trend("A") + season("A")),
    "Holt-Winters_damped" = ETS(box_cox(average, lambda) ~ error("A") + trend("Ad") + season("A")), 
    "ETS_auto" = ETS(box_cox(average, lambda)),
     arima = ARIMA(box_cox(average, lambda), stepwise = FALSE, approx = FALSE),
    "Theta_multiplicative" = THETA(box_cox(average, lambda) ~ season(method = "multiplicative")),
    "Theta_additive" = THETA(box_cox(average, lambda) ~ season(method = "additive")),
    "Prophet_A" = prophet(box_cox(average, lambda) ~ season(period = 12, order = 2,
                                                      type = "additive")),
    "Prophet_M" = prophet(box_cox(average, lambda) ~ season(period = 12, order = 2,
                                                      type = "multiplicative")),
     N_Net = NNETAR(box_cox(average, lambda)),
     Mean = MEAN(average),
    `Naïve` = NAIVE(average),
    Drift = NAIVE(average ~ drift()),
    `Seasonal naïve` = SNAIVE(average)
  ) |>
  mutate(comb5 = (Prophet_M + Prophet_A) / 2) %>% 
  mutate(comb6 = (Prophet_M + arima) / 2) %>% 
  mutate(comb7 = (arima + Prophet_A) / 2) %>%  
  mutate(comb8 = (Prophet_M + Prophet_A + arima) / 3)

my_fc <- my_fit |> 
  forecast(h = h)

# test the forecast accuracy on the test set.
accuracy(my_fc, mydata) |>
  arrange(RMSE) 

```

Initially, the top three models by RMSE from the chosen shortlist of advanced and basic techniques were: Prophet_A, Prophet_M, and arima (specifically ARIMA(1,1,1)(0,1,1)[12]). Accordingly, the above model comparison also features four possible multipart combinations of these top three models. In terms of root mean squared error (RMSE), the most effective two models of the combinations were comb5 (comprised of Prophet_M and Prophet_A), and comb8 (Prophet_A, Prophet_M, and arima). With very little difference in RMSE, the top models were Prophet_A, Prophet_M, comb5, and comb8. Each iteration of running the code analyzing model accuracy produced different exact RMSE results, but each of these models possessed an RMSE of approximately 3. The best simple forecasting method, Drift, faded in contrast, with the four superior models by RMSE typically less than half of Drift's own RMSE.


```{r}
# Residual properties
my_fit |>
  select(arima) |> 
  gg_tsresiduals() +
  ggtitle("Arima Residuals")


# The Ljung-Box test
my_fit |>
  select(arima) |>
  augment() |>
  features(.innov, ljung_box, lag = 20)
```

Before viewing the results the Ljung-Box test, our candidate models of the shortlist evaluated above largely fits many of the guidelines for reliable forecasting. Crucially, each residual plot centers around 0 with a near-Normal distribution. To the naked eye, the residual time plots of the most recent models do not possess concerning levels of heteroskedasticity. These features assure forecasters of fairly reliable, stable model estimates, but any failure of the Ljung-Box test by the models with the lowest RSME would signal problems with any potential forecast.

Accordingly, analysis revealed issues with our top-ranking models. Both Prophet_A and Prophet_M failed the Ljung-Box test with low p-values approaching 0, indicating pattern within the residuals unaccounted for by each model. Combination models featuring each of Prophet_A and Prophet_M exhibited similar problems. Proceeding down the list of singular and ensemble models by accuracy, the first to contain residual behavior close enough to white noise was "arima", with a Ljung-Box p-value of 0.6591.

Furthermore, of arima itself and models containing lower RMSE values, only the ARIMA model exhibits a remotely acceptable of its ACF values within the target threshold. The amount of ACF values above this threshold in other candidate models generally hovers around the 10% range or above, which dissuades from their potential selection. The ACF plots and Ljung-Box test results together indicate that we maintain more confidence in the ARIMA model lacking residual patterns that render model application difficult for the future.  

Despite the ARIMA model's lesser accuracy with an RMSE of 4.422 when compared to several of the other models we attempt to choose from, its assurance against residual autocorrelation marks it as a more reliable forecaster. For reference, Prophet_A, Prophet_M, comb5, and comb8 had RMSE values about 30% lower for each iteration of this analysis. We still prefer the ARIMA model, despite sacrificing some predictive accuracy, for its stronger grasp of ideal white noise residuals that imply a capture of most relevant data information when crafting the model. This model still represents an improvement over simple forecasting techniques, less than half of Drift's RMSE as the best simple model for our data. Thus, the ARIMA model will comprise the next forecasting step.


```{r}
# Fit the model of your choice
my_fit <- mydata |> # notice you are using the complete data here, not the training set
  model(
    "arima" = ARIMA(box_cox(average, lambda), stepwise = FALSE, approx = FALSE) # your preferred method could differ.
)


# Forecast for 10 periods ahead.
my_fit |> 
  forecast(h = 120)

#plotted data starting in 1990 to view the forecast better
my_fit %>% 
  forecast(h = 120, level = c(80, 95)) %>% 
  autoplot(mydata %>% 
             filter(date > as.Date("1989-12-01"))) 

```

The new forecast 10 years into the future, from the ARIMA model, projects a continuation of our seasonality and positive trend, but a higher rate of increase than ever previously seen since the data's beginning in 1958. The next section of this report explores the results and their implications in further detail.

# Results

The rate of increase in Mauna Loa CO2 levels is itself set to increase, with a peak CO2 levels 3.41 ppm higher than in 2024 to 430.32 in Year 1 (2025) of our forecast, and then increasing from there to 432.87 in Year 2. In corresponding fashion, our forecast projects new record Mauna Loa CO2 levels with each passing year and claims we expect values above 450 ppm within 10 years (specifically a peak of 453.32 in May 2034). This projected increase in annual peak of over 20 ppm in one decade outranks the average rate of increase from the 66 recorded years of Mauna Loa readings, which averaged closer to an increase of 17 ppm per decade. 

This numerical CO2 forecast signals the prospect of a steeper rate of temperature increases, frequency and severity of natural disasters, and other climate disruptions. Considering the global goal to curb the worst effects of climate change, our findings cast doubt on the efficacy of current measures to alleviate these damages. Costs incurred to both the public and private sectors, too, will likely increase due to climate change damages.

However, as our forecast horizon increases our level of certainty also decreases. Although we maintain confidence of at least 80% in a continued increase, the true long-term rate may prove worse or slightly less bleak than these point estimates. In the worst-case scenario, with CO2 levels at the highest end of our 80% confidence level, the forecast develops a more exponential quality where the monthly average of CO2 readings looks set to spiral out of control. On the opposite end, if true future values mirror the lowest points of the 80% confidence level, the positive trend appears far more linear and manageable, though still carrying the promise of further environmental changes. 

Regardless of exact accuracy, our forecast implies that stakeholders, namely businesses whose practices require shifts due to climate change, need to adjust their practices to account for the exponentially increasing effects of climate change occurring partially due to the increase of atmospheric CO2 levels. While climate change is demonstrable and businesses have adapted, businesses need to increase their rate of adaptation to match the increasing rate of atmospheric CO2's rise. Current processes in many industries may prove insufficient, since a rate of transition that succeeded in the 2010s decade could now be too slow to match the changes in the atmosphere.  

# Discussion and Conclusion

The findings in this report justify the growing feelings of alarm concerning climate change worldwide. The fact that even conservative estimates at the lower end of our 80% confidence level for a horizon of 10 years place CO2 levels at a clear rate of increase removes much of the autonomy of businesses regarding climate change policies. In affected industries, the only choice now becomes how quickly to adapt and what short-term sacrifices to make to avoid catastrophic blows to profits. It is impossible to depend on the status quo in a violently changing world, even if CO2 levels do not quite surpass 450 ppm by 2034 from a current all-time high of 426.91 as this study predicts. Even if overly pessimistic, this CO2 forecast still certainly warns against avoiding impending climate-based difficulties. 

Potential limitations of this study include the lack of secondary variables with which to draw conclusions. Although CO2 is a large part of the climate change discussion, other greenhouse gases also drive up global temperatures such as methane. Theoretically, a large change in the levels of one could offset the other. Many sources including NOAA, the source of this analysis, also carry data for other greenhouse gases, so a logical next step would include a more comprehensive forecast of greenhouse gases to deliver firmer contextual results. 

Likewise, a stronger study could enlist data from another observatory elsewhere in the world to better support its findings. Although Mauna Loa's observatory is reputable in its field, diversification of our sources adds certainty to any drawn conclusions. A continuation of more thorough studies into climate change factors and effects will help prepare businesses, governments, and individual citizens alike in understanding and adjusting to the changes taking place in the natural world. If more natural disasters and accrued damages remain inevitable, perhaps with superior forecasting their effects may be minimized.



# References

Betts, Richard A.; Jones, Chris D.; Keeling, Ralph; Knight, Jeff R., Pope, James O., Sandford,      Caroline. Mauna Loa carbon dioxide forecast for 2025. Met Office.
     https://www.metoffice.gov.uk/research/climate/seasonal-to-decadal/long-range/forecasts/co2       -forecast-for-2025#:~:text=We%20forecast%20the%20annual%20average,of%20Oceanography%2C%20      UC%20San%20Diego.
     
Trends in CO2, CH4, N20, SF6. NOAA. 
     https://gml.noaa.gov/ccgg/trends/data.html